Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods

### 2022-04-22 inst/surfing_jsa_rebuttal_referee2.tex

The paper does a good job of outlining the basics of surfing in terms
of competition rules and structure, and frames up the problem at
hand.  However, the paper itself feels rather light on content.  The
authors jump almost immediately to the results section and don't go
into very much detail in terms of how or why Bradley-Terry models are
used, other than the fact that they are used.

\textcolor{blue}{We have re-read our manuscript and indeed
  Bradley-Terry (BT) does make a rather abrupt entrance.  However, we
  do describe and justify and motivate the use of BT in section three.
  On reflection we believe we can address this comment by rearranging
  sections 2 and 3: this gives a gentler and more appropriate
  introduction to BT and indeed is a more appropriate structure for a
  manuscript.  In line with this, we have changed the section heading
  from ``previous related work'' to ``Previous statistical analysis of
  competitive surfing'' which is a better description, we feel.}

The authors are guilty
of referring to various numerical quantities throughout the paper,
without expanding on how these values were calculated, and what
implications the values have. The paper would benefit from more
explicit equations throughout; on several occasions the authors refer
to modifying or changing their likelihood function to suit their
problem at hand, but it is difficult for the reader (or at least
myself!) to intuitively understand what this modified function would
look like, without writing it down or thinking hard about
it. Throughout the paper I found I had to do a lot of the mathemati
cal and abstract thinking myself, which would have been made much
easier with more equations and explicit definitions and derivations of
parameters, likelihood functions, and important quantities that are
referred to in Sections 5, 6, and 7.

\textcolor{blue}{These comments are entirely reasonable and, on
  re-reading the manuscript, we agree that we could have done a much
  better job describing the mathematical and statistical basis for our
  conclusions, especially the motivation for the likelihood equations.
  In the resubmission we have added a large amount of explanation that
  hopefully makes our reasoning more clear.\\ \\
  However, we observe that {\em all} our calculations are available
  for public inspection on gihub at our repo {\tt RobinHankin/hyper2}.
  In particular, file {\tt inst/surfing.Rmd} can reproduce all our
  findings.  However, the nitty-gritty detail of our R idiom is, we
  suggest, not really suitable for a JSA manuscript.  We believe that
  our enhanced reporting in the revision is a good compromise between
  not enough detail and too much.
  \begin{itemize}
  \item We have provided more detail (mathematical and computational)
    for the Fisher information argument for retaining the data with
    three surfers, as opposed to retaining just the pairs
  \item Added detail on the computational infrastructure for locating
    maximum likelihood estimates
  \item Referenced the computational method [{\tt equalp.test()} for
    testing the null of equal competitive abilities
  \item Introduced a new table that explicitly sets out the likelihood
    function for the reified entity Brazilian beachtype reified entity
  \end{itemize}
}

### 2022-04-22 inst/surfing_jsa.tex
further work on issue #129

\documentclass{article}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[pdftex]{hyperref}

@@

We can directly calculate the determinant of the observed information
matrix for the likelihood function in
equation~\ref{complete_likelihood_function} at about $4.3\times
10^{71}$, as compared with~$2.9\times 10^{69}$ for the pairs
considered in isolation, a ratio of about $146.4$.  This ratio shows
that there is considerable information loss in considering only the
paired comparisons.  We are thus motivated to consider the whole
dataset, not just the part which involves paired comparisons.

@@

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
  hyper2} software~\citep{hankin2017} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  The Method of Support \citep{edwards1992}
rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
There is thus strong evidence to suggest that the competitors are
indeed of different skills and the result actually reflects genuine
differences in abilities, rather than the result of random chance.
Alternatively, we may view the dataset as being sufficiently
informative to reveal actual differences in surfing ability.  Observe
that the competitors' points (that is, the final column of
table~\ref{resultstable}) cannot be used for this type of
probabilistic analysis: the points awarded are arbitrary and changing
the points system would change the points scored but not affect the
competitors' behaviour.

@@

We may ask whether the top two officially ranked competitors, Medina
and Ferreira, have different strengths.  To test the null of equal
strength, we maximize the likelihood subject to the null, and compare
with the global maximum likelihood~\citep{hankin2010}.  This gives us a
support difference of about 0.05, well short of Edwards's 2 units of
support criterion.  Alternatively, we could observe that the
$-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the
tail region of its asymptotic $\chi^2_1$ distribution.  We may thus
assert that there is no strong evidence for Ferreira's actually having
a competitive advantage over Medina.  We may similarly wonder whether
Medina is equal in strength to Wilson; here we reject the null with
support difference of about 2.33, corresponding to a $p$-value of
$0.031$.

@@

We now assess the null that Brazilians perform equally well on all
types of wave types.  To that end, we create a reified surfer,
``Brazilian wavetype", whose strength helps all Brazilian surfers
when surfing on their (putative) preferred wave type, an analogue of
the ``home ground monster" who helps the home team as per
equation~\ref{home_ground_advantage}.  Here we modify the likelihood
function only when a single Brazilian is in the water, thus avoiding
possible interaction with the compatriot issue discussed in
section~\ref{compatriot}.  The null would be that this reified entity
has zero strength.  We fail to reject this null (differential support
= 0.035; $p=0.79$) and conclude that there is no evidence to suggest
that Brazilians' performance is better on their preferred wave types
than one would expect by chance.  We may go further using the method
of support: not only is the maximum likelihood estimate for Brazilian
wavetype very small (about 0.002), but in addition we may reject the
the proposal that the strength of this entity exceeds about 0.025
(Figure~\ref{brazilian_wavetype_support}).

@@

>>>>

\documentclass{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage[pdftex]{hyperref}

@@

To answer this, we directly calculate the observed information matrix
for the log-likelihood function in
equation~\ref{complete_likelihood_function}.  Observing that
differentiation is linear, we may calculate second derivatives by
summing individual terms of the support function and further observing
that, for a typical term of the form $\log\left(\sum_{k\in K}
p_k\right)^n$ [where the sum is over some subset $K$ of strengths] we
have

\begin{equation}
  \frac{\partial^2}{\partial p_i\partial p_j}
  \log\left(\sum p_k\right)^n=
  \begin{cases}
    -n/\left(p_i+p_j+\sum p_k\right)^2\qquad\mbox{if $i,j\in K$}\\
    0\qquad\mbox{otherwise.}
    \end{cases}
\end{equation}

The {\tt hyper2} package includes function {\tt hessian()} which
returns this matrix for any given support function.  For
equation~\ref{complete_likelihood_function} we find that the
determinant is about $4.3\times 10^{71}$.  Compare this with
~$2.9\times 10^{69}$ for the pairs considered in isolation, a ratio of
about $146.4$.  This ratio shows that there is considerable
information loss in considering only the paired comparisons.  We are
thus motivated to consider the whole dataset, not just the part which
involves paired comparisons.

@@

For any likelihood function, one important diagnostic is the maximum
likelihood estimate.  The {\tt hyper2} package facilitates the finding
of maximum likelihood estimates using the bespoke {\tt maxp()}
function, which uses standard optimization techniques.  Derivatives
are available which ensures rapid convergence to the evaluate.

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
  hyper2} software~\citep{hankin2017} may be used to formally assess
one plausible null: that the competitors all have equal skill, that is
$H_0\colon p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  This test is performed by function {\tt
  equalp.test()} in the package which uses the method of support
\citep{edwards1992}.  The test rejects the null (in favour of the
evaluate) at $p=4\times 10^{-5}$.  There is thus strong evidence to
suggest that the competitors are indeed of different skills and the
result actually reflects genuine differences in abilities, rather than
the result of random chance.  Alternatively, we may view the dataset
as being sufficiently informative to reveal actual differences in
surfing ability.  Observe that the competitors' points (that is, the
final column of table~\ref{resultstable}) cannot be used for this type
of probabilistic analysis: the points awarded are arbitrary and
changing the points system would change the points scored but not
affect the competitors' behaviour.

@@

We may ask whether the top two officially ranked competitors, Medina
and Ferreira, have different strengths.  To test the null of equal
strength, we maximize the likelihood subject to the null, and compare
with the global maximum likelihood~\citep{hankin2010}; package idiom
is to use {\tt equalp.test()} which reports its findings in a
standardised format.  This gives us a support difference of about
0.05, well short of Edwards's 2 units of support criterion.
Alternatively, we could observe that the
$-2\log\left(\frac{\Lambda}{\Lambda_0}\right)\simeq 0.1$ is not in the
tail region of its asymptotic $\chi^2_1$ distribution.  We may thus
assert that there is no strong evidence for Ferreira's actually having
a competitive advantage over Medina.  We may similarly wonder whether
Medina is equal in strength to Wilson; here we reject the null with
support difference of about 2.33, corresponding to a $p$-value of
$0.031$.

@@

We now assess the null that Brazilians perform equally well on all
types of wave types.  To that end, we create a reified surfer,
``Brazilian wavetype", whose strength helps all Brazilian surfers
when surfing on their (putative) preferred wave type, an analogue of
the ``home ground monster" who helps the home team as per
equation~\ref{home_ground_advantage}.  Here we modify the likelihood
function only when a single Brazilian is in the water, thus avoiding
possible interaction with the compatriot issue discussed in
section~\ref{compatriot}.

To illustrate the concept, we show a simple likelihood function in
Table~\ref{probabilitybrazilian}.  The null would be that this reified
entity has zero strength.  We fail to reject this null (differential
support = 0.035; $p=0.79$) and conclude that there is no evidence to
suggest that Brazilians' performance is better on their preferred wave
types than one would expect by chance.  We may go further using the
method of support: not only is the maximum likelihood estimate for
Brazilian wavetype very small (about 0.002), but in addition we may
reject the the proposal that the strength of this entity exceeds about
0.025 (Figure~\ref{brazilian_wavetype_support}).

@@

\begin{table}[h]
\begin{tabular}{l|l|l}
wave type  & $\operatorname{Prob}(p_1\succ p_2)$ & $\operatorname{Prob}(p_2\succ p_1)$\\ \hline
beach & $\frac{p_1+B}{p_1+p_2+B}$ & $\frac{p_2}{p_1+p_2+B}$\\
point & $\frac{p_1+B}{p_1+p_2+B}$ & $\frac{p_2}{p_1+p_2+B}$\\
reef  & $\frac{p_1\vphantom{B}}{p_1+p_2}$ & $\frac{p_2}{p_1+p_2}$\\
\end{tabular}
\caption{Likelihood function for results between surfer 1 (Brazilian)
  and surfer 2 (non-Brazilian) with \label{probabilitybrazilian} Bradley-Terry
  strengths $p_1,p_2$ respectively, and a ``Brazilian wavetype" entity with
  strength $B$ that quantifies the enhanced Brazilian performance on reef wave types}
\end{table}
